---
title: "Simulation and signal drift correction with pseudoDrift"
author: "Jonas Rodriguez"
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
        toc: true
        toc_depth: 2  
        number_sections: true  
        toc_float: true
bibliography: [pseudoDrift.bib]
csl: analytical-chemistry.csl
vignette: >
    %\VignetteIndexEntry{pseudoDrift Simulation and Correction with pseudoQCs}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

We will demonstrate the functionality of pseudoDrift by data simulation followed by signal drift correction. To get started, make sure you have the most recent development version from [GitHub](https://github.com/) installed. Throughout this document, QC is meant to describe quality control or replicate samples of the same type. These are typically included in metabolomics studies as a way to identify and correct technical errors or other anomalies which may occur during the data acquisition phase of an experiment.

```{r, eval=TRUE, message=FALSE}
# install.packages("remotes")
# remotes::install_github("jrod55/pseudoDrift")

library(pseudoDrift)
library(tidyverse)
library(data.table)
```

# Data simulation
To simulate data, first we need an indexed and validated .sdf file to pull the metadata associated with the compound we are interested in. We are using a small .sdf file which is included as part of this package, but to construct your own, simply run the command below. The `sdf2index` function uses a few ChemmineR R package [@R-ChemmineR] functions to create an indexed file so that the entire .sdf file doesn't need to be loaded into memory.

```{r, eval=FALSE}
# sdf2index(sdf_input = "/path_to_your/desired.sdf", out_name = "Index.xls") # Example
sdf2index(sdf_input = system.file("extdata", "test.sdf", package = "pseudoDrift"), out_name = "Index.xls") # Can run, but not necessary for this vignette
```

The `simulate_data` function takes as input a compound name (which should be present in the .sdf file) or a db ID (also should be present in the .sdf file) and simulates peak matrices using m/z values for the queried compound. We can also adjust the number of batches `nbatch`, number of samples per batch `nsamps_per_batch`, and the frequency of QC samples `QC_freq` during the simulation. Lastly, the `seed` can be changed to generate a new simulation.

```{r, eval=TRUE}
sim1 = simulate_data(compound_names = "tricin",
                     xls_file_name = system.file("extdata", "Index.xls", package = "pseudoDrift"),
                     valid_sdf_file = system.file("extdata", "valid-test.sdf", package = "pseudoDrift"),
                     nbatch = 3,
                     nsamps_per_batch = c(25, 50, 100),
                     QC_freq = c(5,15,20),
                     seed = 123)
```

The code above generates simulated data for each instance of 'tricin' found in the .sdf file. The sim1 object is a nested tibble which contains all metadata available for the compound, along with the simulated data (sim_mat) and four signal drift and batch effect types: 1) monotonic (t1_sim_mat) , 2) batch-to-batch block effect (t2_sim_mat) , 3) random (t3_sim_mat), 4) monotonic coupled with batch-to-batch effect (t4_sim_mat).

**Column names of the nested tibble returned by** `simulate_data`
```{r, eval=TRUE}
names(sim1)
```

From the nested tibble above, we can isolate an individual row using the compound id from in the first column then pull the corresponding simulated data using list indexing syntax. In the code below, we are going to isolate the simulated data from the first id: FIO00738 then visualize the signal drift trends simulated. 

```{r, eval=TRUE}
sim1_sub = sim1 %>% 
  filter(id=="FIO00738")

sim_dat = sim1_sub$sim_mat[[1]]         # simulated no effects
mono = sim1_sub$t1_sim_mat[[1]]         # monotonic
b2b = sim1_sub$t2_sim_mat[[1]]          # batch-to-batch
rando = sim1_sub$t3_sim_mat[[1]]        # random
mono_b2b = sim1_sub$t4_sim_mat[[1]]     # monotonic and batch-to-batch
```

Plot and highlight the QC samples

```{r, fig.height=5, fig.width=6.5, eval=TRUE}
# small plotting function
plt_fun = function(x,dat_lab){
  qc_samps = x %>% filter(sample=="QC")
  plt = ggplot(x, aes(batch_index, area))+
    geom_point()+
  geom_point(data = qc_samps, aes(color=sample))+
  geom_line(data = qc_samps, aes(color=sample))+
  facet_grid(cols = vars(batch), scales = "free")+
  labs(title = paste0(dat_lab))
  return(plt)
}

plt_fun(sim_dat, "Simulated")
plt_fun(mono, "Monotonic")
plt_fun(b2b, "Batch-to-batch")
plt_fun(rando, "Random")
plt_fun(mono_b2b, "Monotonic and batch-to-batch")

```

# Signal drift correction
We will now focus only on the type 4 signal drift and batch effect simulation. Primarily because this is what is most typically observed in real datasets. Suppose we ran a large metabolomics experiment such as the one simulated in the previous section. Only now, we assume that QC samples were run in batch 3 alone. Using the data from batch 3 as a training batch, we estimate three parameters (`test.breaks`, `test.window`, `test.index`) from the non-QC samples, to predict the signal drift trends in the remaining batches with what we call 'pseudoQC' samples.

```{r, eval=TRUE}
train.batch = "B3"         # Batch to be used as training set
test.breaks = seq(10,12,1) # Test breaking up the training batch into 10 to 12 equally sized sections
test.window = seq(1,9,2)  # Test taking the median every 1 (reduces to the mean per test.break), to 9 samples. This sequence of numbers should be all odd.
test.index = seq(0,5,5)   # Test offsetting the injection order of pseudoQC samples estimated
criteria = "MSE"           # This can be one of: "RSD" (relative standard deviation), "MSE" (mean squared error), or "TSS" (total sum of squares). This is the criteria to be minimized.
n.cores = 10               # Number of cores to use if your machine has the ability to multi-thread processes.


sdc_out = pseudo_sdc(df = mono_b2b,
                     train.batch = train.batch,
                     test.breaks = test.breaks,
                     test.window = test.window,
                     test.index = test.index,
                     criteria = criteria,
                     qc.label = "QC",
                     n.cores = n.cores,
                     min.qc = min(test.breaks))

```

Internally, the `psuedo_sdc` function fits all possible regression splines with the derived 'pseudoQC' samples calculated from the data in the training batch. These regressions are fit using the Quality Control-Robust Spline Correction `QCRSC` function from the pmp R package [@R-pmp] and finally the optimal combination of input parameters are applied to the training data to estimate 'pseudoQC' samples and apply the correction between batches and for signal drift. \

The figure below illustrates the raw data in (A) and the signal drift and batch corrected data using pseudo_sdc (B). 

```{r, eval=TRUE, echo = FALSE, fig.height=8.5, fig.width=6.5}
# slightly larger plotting function
plt_fun1 = function(x,train.batch){
  metab = paste0(unique(x$df$compound),"_SIMULATED")
  x1 = x$df_pseudoQC %>% 
    group_by(batch) %>% 
    mutate(index = 1:n(),
           class = factor(class, levels = c("QC", "Pseudo_QC", "Sample")))
  
  qcs1 = x1 %>% 
    filter(class%in%c("QC","Pseudo_QC")) %>% 
    mutate(sample = class)
  
  plt1 = ggplot(x1, aes(index, area))+
    geom_point()+
    geom_point(data = qcs1, aes(color=sample))+
    geom_line(data = qcs1, aes(color=sample))+
    facet_grid(cols = vars(batch), scales = "free")+
    labs(title = paste0(metab," raw data using ",train.batch, " data to train pseudo-QC"))
  
  legend = cowplot::get_legend(plt1)
  
  x2 = x$df_pseudoQC_corrected %>% 
    group_by(batch) %>% 
    mutate(index = 1:n(),
           class = factor(class, levels = c("QC", "Pseudo_QC", "Sample")))
  
  qcs2 = x2 %>% 
    filter(class%in%c("QC","Pseudo_QC")) %>% 
    mutate(sample = class)
  
  plt2 = ggplot(x2, aes(index, area_corrected))+
    geom_point()+
    geom_point(data = qcs2, aes(color=sample))+
    geom_line(data = qcs2, aes(color=sample))+
    facet_grid(cols = vars(batch), scales = "free")+
    labs(title = paste0(metab," pseudo-QC corrected data"))
  
  cplt = cowplot::plot_grid(plt1, plt2, ncol = 1, labels = c("A","B"))
  return(cplt)
}

plt_fun1(sdc_out,train.batch)
```

## Comparing correction with originally simulated data

```{r eval=TRUE, fig.height=8.5, fig.width=6.5}
cor_dat = sdc_out$df_pseudoQC_corrected %>% 
  select(name:area_corrected, class) %>% 
  left_join(., sim_dat) %>% 
  drop_na(area)

cor_dat %>% 
  ggplot(., aes(area_corrected, area, color = batch))+
  geom_point()+
  labs(x = "area_corrected_with_pseudoQC",
       y = "original_area_simulated_without_any_effects")

  
library(caret)
fc = trainControl(method = "cv", number = 10)
fit = train(area ~ area_corrected, 
             data = cor_dat, 
             method = "lm", 
             trControl = fc)

plot(fit$finalModel)
```

# Session information

```{r eval=TRUE}
sessionInfo()
```


